Accelerating parameter inference with graphics processing units

### 2019-02-03 paper/paper.tex
paper.tex: draft almost ready for P&P circulation

\title{Accelerating parameter inference with graphics processing units}
\author{D. Wysocki}
\affiliation{\RIT}
\author{Yao-Lung L. Fang}
%\author{Yao-Lung L. Fang (方耀龍)}
\affiliation{Computational Science Initiative, Brookhaven National Laboratory, Upton, NY 11973, USA}
\affiliation{National Synchrotron Light Source II, Brookhaven National Laboratory, Upton, NY 11973, USA}
\author{R. O'Shaughnessy}
\affiliation{\RIT}
\author{Jake?}
\affiliation{\RIT}
\begin{abstract}
Gravitational wave Bayesian parameter inference involves repeated comparisons of GW data to generic candidate predictions.
Even with algorithmically efficient methods like RIFT or reduced-order quadrature, the time needed to perform these
calculations and overall computational cost can be significant compared to the minutes to hours needed to achieve the
goals of low-latency multimessenger astronomy.  By translating some elements of the RIFT algorithm to operate on
graphics processing units (GPU), we demonstrate substantial performance improvements, enabling dramatically reduced
overall cost and latency. 
\end{abstract}
\maketitle

@@

\begin{figure*}
% COMMAND
% plot_posterior_corner.py --posterior-file posterior_samples-2.dat --posterior-file posterior_samples-1.dat --parameter m1 --parameter m2  --plot-1d-extra --ci-list  '[0.9]' --truth-file /home/oshaughn/unixhome/Projects/LIGO-ILE-Applications/ILE-GPU/ILE-GPU-Paper/demos/mdc.xml.gz  --composite-file all.net --quantiles None  --posterior-label 2 --posterior-label 1 --use-legend
\includegraphics[width=0.45\textwidth]{figures/bns_zerospin_m1_m2.png}
\includegraphics[width=0.45\textwidth]{figures/bns_zerospin_mc_cum.png}
\includegraphics[width=0.45\textwidth]{figures/bns_zerospin_lnL_meanVar.png}
\includegraphics[width=0.45\textwidth]{figures/bns_zerospin_lnL_converge.png}
\caption{\label{fig:BBH:MultiIterate}\textbf{Convergence of BBH analysis: Zero spin}: Results for marginal posterior distributions
  of our fiducial synthetic neutron star.  Solid contours show credible intervals; solid one-dimensional distributions
  show marginal CDFs and PDFs for the corresponding variable; and colored points indicate the location $\lambda$ and
  value of the underlying marginalized likelihood evaluations.   The corresponding dotted curves show an analysis using
  only the $m=\pm 2$ modes \editremark{perform both} \emph{Left panel } Posterior distribution
  over  $\mc$ and
  $\delta=(m_1-m_2)/M$.    \emph{Right panel}: Marginal 1d CDFs of $\mc$, showing convergence.
\emph{Bottom left}: Mean and variance of  $\ln{\cal L}_{\rm marg}$ on the evaluation points,  showing that after the
first iteration the
candidate points are consistent with the posterior (i.e., no proposed point has very low $\ln {\cal L}_{\rm marg}$).
\emph{Bottom right panel}: $\ln {\cal L}_{\rm marg}$ versus iteration number.  As systematic fitting error dominates our
error budget early on, the Monte Carlo error is only shown for the last iteration.
}
\end{figure*}

Modest-amplitude short-duration binary black holes empirically constitute the most frequent detection candidates
for current ground-based GW observatories \cite{LIGO-O2-Catalog}.    Because of their brevity and hence broad posterior,
the RIFT code's interpolation-based method converges  rapidly.  Combined with the low cost of each iteration,
these sources require an exceptionally low committment of resources and can be performed in extremely low latency, as
desired.
To demonstrate this, we use the GPU-accelerated code with $N_{\rm eval}=20$, analyzing data with $f_{\rm
  sample}=4096\unit{Hz}$.    We use 100 points in the first iteration, in a coarse mass  grid with astrophysically significant
scale, followed by 20 points in each subsequent iteration.  Figure \ref{fig:BBH:MultiIterate} shows our results.   The
final posterior is already well-explored with the initial gridpoints, and converged by the second iteration.  
Conversely, on average all iterations required roughly 45 seconds per $\lambda$ to evaluate their grid.  This
configuration therefore converged within 30 minutes.    We can achieve  smaller turnaround time for an otherwise identical analysis by adjusting $N_{\rm eval}$ appropriate to the available hardware.
%
While this low-dimensional problem does not capture all  fitting and automation challenges associated with high-dimensional
 fully precessing binaries, it does capture the low cost and rapid response possible with RIFT.

@@

\begin{figure*}
% COMMAND
% plot_posterior_corner.py --posterior-file posterior_samples-4.dat --posterior-file posterior_samples-3.dat --posterior-file posterior_samples-2.dat --posterior-file posterior_samples-1.dat --parameter mc --parameter delta_mc --mc-range  '[1.19655,1.19682]' --plot-1d-extra --ci-list  '[0.9]' --truth-file /home/oshaughn/unixhome/Projects/LIGO-ILE-Applications/ILE-GPU/ILE-GPU-Paper/demos/mdc_bns.xml.gz  --composite-file all.net --quantiles None --posterior-label 4 --posterior-label 3 --posterior-label 2 --posterior-label 1 --use-legend
\includegraphics[width=0.45\textwidth]{figures/bns_zerospin_corner_mc_delta_mc.png}
\includegraphics[width=0.45\textwidth]{figures/bns_zerospin_mc_cum.png}
\includegraphics[width=0.45\textwidth]{figures/bns_zerospin_lnL_meanVar.png}
\includegraphics[width=0.45\textwidth]{figures/bns_zerospin_lnL_converge.png}
\caption{\label{fig:BNS:MultiIterate}\textbf{Convergence of BNS analysis: Zero spin}: Results for marginal posterior distributions
  of our fiducial synthetic neutron star.  Solid contours show credible intervals; solid one-dimensional distributions
  show marginal CDFs and PDFs for the corresponding variable; and colored points indicate the location $\lambda$ and
  value of the underlying marginalized likelihood evaluations.   The corresponding dotted curves show an analysis using
  only the $m=\pm 2$ modes \editremark{perform both} \emph{Left panel } Posterior distribution
  over  $\mc$ and
  $\delta=(m_1-m_2)/M$.    \emph{Right panel}: Marginal 1d CDFs of $\mc$, showing convergence.
\emph{Bottom left}: Mean and variance of  $\ln{\cal L}_{\rm marg}$ on the evaluation points,  showing that after the
first iteration the
candidate points are consistent with the posterior (i.e., no proposed point has very low $\ln {\cal L}_{\rm marg}$).
\emph{Bottom right panel}: $\ln {\cal L}_{\rm marg}$ versus iteration number.  As systematic fitting error dominates our
error budget early on, the Monte Carlo error is only shown for the last iteration.
}
\end{figure*}

\begin{figure*}
\caption{\label{fig:BNS:Comparison}\textbf{Comparison of GPU, CPU analysis}: \editremark{worthwhile}
}
\end{figure*}

Significant-amplitude  binary neutron star mergers are the most important scenario for rapid parameter inference, as low
latency can enable multimessenger followup  \cite{LIGO-O2-Catalog}.    
% 
In our second test we compare two workflows, one with the original embarassingly-parallel RIFT code using $N_{\rm
  eval}=1$ and one with the
``batched'' GPU-accelerated code with $N_{\rm eval}=20$, analyzing data with $f_{\rm sample}=4096\unit{Hz}$.    
% Code configuration
In this example, we use $100$ initial points spread over the two mass dimensions $\mc_z,\delta$, adding $20$ evaluations per
iteration.   We
choose this simple low-dimension, small-size, and slowly-converging configuration to facilitate visualization of the grid, posterior, and convergence.     The initial coarse grid covers a region $\mc_z\in[1.1962,1.1970] M_\odot$ and $\delta \in [0,0.25]$,  insuring the posterior was smaller than our initial
coarse grid spacing.     The top  panels of Figure  \ref{fig:BNS:MultiIterate} show posterior disributions derived
from the first several RIFT iterations, while the bottom panels show convergence diagnostics.
Figure \ref{fig:BNS:Comparison} compares results from the GPU and CPU codes on \editremark{identical test points: the
  initial grid}.\footnote{This comparison of $\ln {\cal L}_{\rm marg}$ is limited by Monte Carlo integration error and
  is provided  as a pedagogically convenient sanity check.  When comparing 
  $\ln {\cal L}_{\rm margT}$ results on identical data and parameters, the GPU and CPU implementations agree to
  significantly higher precision.}  The two results agree.
Because our analysis uses a post-Newtonian model to interpret a signal generated with SEOBNRv4, the peak posterior
density is slightly  offset
 from the synthetic signal's parameters

@@

As a concrete demonstration of a low-latency analysis using the GPU-accelerated code, we reanalyze our fiducial binary
neutron star signal, accounting for the possibility of nonzero (aligned) neutron star spins
$\chi_{i,z}\in[-0.05,0.05]$.  
Because  our fiducial post-Newtonian model (TaylorT4) as implemented in \texttt{lalsuite} does not include spin, we
employ both the \texttt{SEOBNRv4\_ROM} and \texttt{SpinTaylorT4} waveform approximations for nonprecessing binaries,
omitting higher-order modes.
%
Figure \ref{fig:BNS:Spin} shows our results.
\editremark{Key feature: runtime: LHO (GPU (c)), 40-ish GPU nodes.  Possibly also overlay SEOBNRv4 analysis as sanity check}

\begin{figure*}
% COMMAND
% plot_posterior_corner.py --posterior-file posterior_samples-4.dat --posterior-file posterior_samples-3.dat --posterior-file posterior_samples-2.dat --posterior-file posterior_samples-1.dat --parameter mc --parameter delta_mc --mc-range  '[1.19655,1.19682]' --plot-1d-extra --ci-list  '[0.9]' --truth-file /home/oshaughn/unixhome/Projects/LIGO-ILE-Applications/ILE-GPU/ILE-GPU-Paper/demos/mdc_bns.xml.gz  --composite-file all.net --quantiles None --posterior-label 4 --posterior-label 3 --posterior-label 2 --posterior-label 1 --use-legend

\caption{\label{fig:BNS:Spin}\textbf{Demonstration of low-latency spinning analysis}:XXX
}
\end{figure*}

>>>>

\title{Accelerating parameter inference with graphics processing units}
\author{D. Wysocki}
\affiliation{\RIT}
\author{Yao-Lung L. Fang}
%\author{Yao-Lung L. Fang (方耀龍)}
\affiliation{Computational Science Initiative, Brookhaven National Laboratory, Upton, NY 11973, USA}
\affiliation{National Synchrotron Light Source II, Brookhaven National Laboratory, Upton, NY 11973, USA}
\author{R. O'Shaughnessy}
\affiliation{\RIT}
\author{Jacob Lange}  % for all the convergence material we work on ... it comes from Jake's practice
\affiliation{\RIT}
\begin{abstract}
Gravitational wave Bayesian parameter inference involves repeated comparisons of GW data to generic candidate predictions.
Even with algorithmically efficient methods like RIFT or reduced-order quadrature, the time needed to perform these
calculations and overall computational cost can be significant compared to the minutes to hours needed to achieve the
goals of low-latency multimessenger astronomy.  By translating some elements of the RIFT algorithm to operate on
graphics processing units (GPU), we demonstrate substantial performance improvements, enabling dramatically reduced
overall cost and latency. 
\end{abstract}
\maketitle

@@

\begin{figure*}
% BASED ON
%   CIT: test_workflow_batch_gpu_lowlatency
% COMMAND
%  plot_posterior_corner.py --posterior-file posterior_samples-4.dat  --posterior-file posterior_samples-3.dat --posterior-file posterior_samples-2.dat --posterior-file posterior_samples-1.dat --parameter m1 --parameter m2  --plot-1d-extra --ci-list  '[0.9]' --truth-file /home/oshaughn/unixhome/Projects/LIGO-ILE-Applications/ILE-GPU/ILE-GPU-Paper/demos/mdc.xml.gz  --composite-file all.net --quantiles None  --posterior-label 4 --posterior-label 3 --posterior-label 2 --posterior-label 1 --use-legend
%plot_posterior_corner.py --posterior-file posterior_samples-2.dat --posterior-file posterior_samples-1.dat --parameter m1 --parameter m2  --plot-1d-extra --ci-list  '[0.9]' --truth-file /home/oshaughn/unixhome/Projects/LIGO-ILE-Applications/ILE-GPU/ILE-GPU-Paper/demos/mdc.xml.gz  --composite-file all.net --quantiles None  --posterior-label 2 --posterior-label 1 --use-legend
\includegraphics[width=0.45\textwidth]{figures/bbh_zerospin_m1_m2.png}
\includegraphics[width=0.45\textwidth]{figures/bbh_zerospin_m1_cum.png}
\includegraphics[width=0.45\textwidth]{figures/bbh_zerospin_lnL_meanVar.png}
\includegraphics[width=0.45\textwidth]{figures/bbh_zerospin_lnL_converge.png}
\caption{\label{fig:BBH:MultiIterate}\textbf{Convergence of BBH analysis: Zero spin}: Results for marginal posterior distributions
  of our fiducial synthetic neutron star.  Solid contours show credible intervals; solid one-dimensional distributions
  show marginal CDFs and PDFs for the corresponding variable; and colored points indicate the location $\lambda$ and
  value of the underlying marginalized likelihood evaluations.   
%The corresponding dotted curves show an analysis using   only the $m=\pm 2$ modes \editremark{perform both} 
\emph{Left panel } Posterior distribution
  over  $\mc$ and
  $\delta=(m_1-m_2)/M$.    \emph{Right panel}: Marginal 1d CDFs of $\mc$, showing convergence.
\emph{Bottom left}: Mean and variance of  $\ln{\cal L}_{\rm marg}$ on the evaluation points,  showing that after the
first iteration the
candidate points are consistent with the posterior (i.e., no proposed point has very low $\ln {\cal L}_{\rm marg}$).
\emph{Bottom right panel}: The estimated evidence $Z = \int d\lambda {\cal L}_{\rm marg}$ versus iteration number.  As systematic fitting error dominates our
error budget, Monte Carlo error is not shown.
}
\end{figure*}

Modest-amplitude short-duration binary black holes empirically constitute the most frequent detection candidates
for current ground-based GW observatories \cite{LIGO-O2-Catalog}.    Because of their brevity and hence broad posterior,
the RIFT code's interpolation-based method converges  rapidly.  Combined with the low cost of each iteration,
these sources require an exceptionally low committment of resources and can be performed in extremely low latency, as
desired.
To demonstrate this, we use the GPU-accelerated code with $N_{\rm eval}=20$, analyzing data with $f_{\rm
  sample}=4096\unit{Hz}$.    We use 100 points in the first iteration, in a very coarse mass  grid (i.e., spacing
comparable to typical astrophysical mass scales), followed by 20 points in each subsequent iteration.  Figure \ref{fig:BBH:MultiIterate} shows our results.   The
final posterior is already well-explored with the initial gridpoints, and converged by the second iteration.  
Conversely, on average all iterations required roughly 45 seconds per $\lambda$ to evaluate their grid on GPU (a) hardware.  This
configuration therefore converged within the 30 minutes needed for the first two iterations.    We can achieve  smaller turnaround time for an otherwise identical analysis by adjusting $N_{\rm eval}$ appropriate to the available hardware.
%
While this low-dimensional problem does not capture all  fitting and automation challenges associated with high-dimensional
 fully precessing binaries, it does capture the low cost and rapid response possible with RIFT.

@@

\begin{figure*}
% COMMAND
% plot_posterior_corner.py --posterior-file posterior_samples-4.dat --posterior-file posterior_samples-3.dat --posterior-file posterior_samples-2.dat --posterior-file posterior_samples-1.dat --parameter mc --parameter delta_mc --mc-range  '[1.19655,1.19682]' --plot-1d-extra --ci-list  '[0.9]' --truth-file /home/oshaughn/unixhome/Projects/LIGO-ILE-Applications/ILE-GPU/ILE-GPU-Paper/demos/mdc_bns.xml.gz  --composite-file all.net --quantiles None --posterior-label 4 --posterior-label 3 --posterior-label 2 --posterior-label 1 --use-legend
\includegraphics[width=0.45\textwidth]{figures/bns_zerospin_corner_mc_delta_mc.png}
\includegraphics[width=0.45\textwidth]{figures/bns_zerospin_mc_cum.png}
\includegraphics[width=0.45\textwidth]{figures/bns_zerospin_lnL_meanVar.png}
\includegraphics[width=0.45\textwidth]{figures/bns_zerospin_lnL_converge.png}
\caption{\label{fig:BNS:MultiIterate}\textbf{Convergence of BNS analysis: Zero spin}: Results for marginal posterior distributions
  of our fiducial synthetic neutron star.  Solid contours show credible intervals; solid one-dimensional distributions
  show marginal CDFs and PDFs for the corresponding variable; and colored points indicate the location $\lambda$ and
  value of the underlying marginalized likelihood evaluations.  
% The corresponding dotted curves show an analysis using   only the $m=\pm 2$ modes \editremark{perform both} 
\emph{Left panel } Posterior distribution
  over  $\mc$ and
  $\delta=(m_1-m_2)/M$.    \emph{Right panel}: Marginal 1d CDFs of $\mc$, showing convergence.
\emph{Bottom left}: Mean and variance of  $\ln{\cal L}_{\rm marg}$ on the evaluation points,  showing that after the
first iteration the
candidate points are consistent with the posterior (i.e., no proposed point has very low $\ln {\cal L}_{\rm marg}$).
\emph{Bottom right panel}: $Z = \int d\lambda {\cal L}_{\rm marg}$ versus iteration number.  As systematic fitting error dominates our
error budget early on, Monte Carlo error is not shown.
}
\end{figure*}

%% \begin{figure*}
%% \caption{\label{fig:BNS:Comparison}\textbf{Comparison of GPU, CPU analysis}: \editremark{worthwhile}
%% }
%% \end{figure*}

Significant-amplitude  binary neutron star mergers are the most important scenario for rapid parameter inference, as low
latency can enable multimessenger followup  \cite{LIGO-O2-Catalog}.    
% 
In our second test we compare two workflows, one with the original embarassingly-parallel RIFT code using $N_{\rm
  eval}=1$ and one with the
``batched'' GPU-accelerated code with $N_{\rm eval}=20$, analyzing data with $f_{\rm sample}=4096\unit{Hz}$.    
% Code configuration
In this example, we use $100$ initial points spread over the two mass dimensions $\mc_z,\delta$, adding $20$ evaluations per
iteration.   We
choose this simple low-dimension, small-size, and slowly-converging configuration to facilitate visualization of the grid, posterior, and convergence.     The initial coarse grid covers a region $\mc_z\in[1.1962,1.1970] M_\odot$ and $\delta \in [0,0.25]$,  insuring the posterior was smaller than our initial
coarse grid spacing.     The top  panels of Figure  \ref{fig:BNS:MultiIterate} show posterior disributions derived
from the first several RIFT iterations, while the bottom panels show convergence diagnostics.
%% Figure \ref{fig:BNS:Comparison} compares results from the GPU and CPU codes on \editremark{identical test points: the
%%   initial grid}.\footnote{This comparison of $\ln {\cal L}_{\rm marg}$ is limited by Monte Carlo integration error and
%%   is provided  as a pedagogically convenient sanity check.  When comparing 
%%   $\ln {\cal L}_{\rm margT}$ results on identical data and parameters, the GPU and CPU implementations agree to
%%   significantly higher precision.}  The two results agree.
Because our analysis uses a post-Newtonian model to interpret a signal generated with SEOBNRv4, the peak posterior
density is slightly  offset
 from the synthetic signal's parameters

@@

As a concrete demonstration of a realistic modest-latency analysis using the GPU-accelerated code, we reanalyze our fiducial binary
neutron star signal, accounting for the possibility of nonzero (aligned) neutron star spins.  Our initial grid consists of 5000 points, spread uniformly across a 4-dimensional
hypercube $\mc_z\in[1.1962,1.1970] M_\odot$, $\delta \in[0,0.25]$ and $\chi_{i,z}\in[-0.05,0.05]$.  Subsequent
iterations use 500 random draws from the estimated posterior samples produced from the previous iteration.  
Because  our fiducial post-Newtonian model (TaylorT4) as implemented in \texttt{lalsuite} does not include spin, we
employ both the \texttt{SEOBNRv4\_ROM} and \texttt{SpinTaylorT4} waveform approximations for nonprecessing binaries,
omitting higher-order modes.   For SEOBNRv4\_ROM, we use 48 GPU (c) enabled nodes; for SpinTaylorT4, we use 100 GPU (a) nodes.
%
Figure \ref{fig:BNS:Spin} shows our results.
%

Due to our self-imposed resource constraints on $N_{\rm eval}=20$ and the number of GPUs, the first iteration is
resource-limited and requires
of order $(5000/20/100)\simeq
2.5$ to $(5000/20/48)\simeq
 5$ times as long as the first iteration of the zero-spin BNS analysis described above.  Each marginalized likelihood
 evaluation in the first requires roughly 60 seconds (SpinTaylorT4) or  (SEOBNRv4ROM), on average.   Subsequent iterations proceed
 in  \editremark{timing}

\begin{figure*}
% COMMAND
% plot_posterior_corner.py --posterior-file posterior_samples-4.dat --posterior-file posterior_samples-3.dat --posterior-file posterior_samples-2.dat --posterior-file posterior_samples-1.dat --parameter mc --parameter delta_mc --mc-range  '[1.19655,1.19682]' --plot-1d-extra --ci-list  '[0.9]' --truth-file /home/oshaughn/unixhome/Projects/LIGO-ILE-Applications/ILE-GPU/ILE-GPU-Paper/demos/mdc_bns.xml.gz  --composite-file all.net --quantiles None --posterior-label 4 --posterior-label 3 --posterior-label 2 --posterior-label 1 --use-legend
\includegraphics[width=0.45\textwidth]{figures/bns_withspin_corner_mc_delta_mc_xi.png}
\includegraphics[width=0.45\textwidth]{figures/bns_withspin_mc_cum.png}
\includegraphics[width=0.45\textwidth]{figures/bns_withspin_lnL_meanVar.png}
\includegraphics[width=0.45\textwidth]{figures/bns_withspin_lnL_converge.png}
\caption{\label{fig:BNS:Spin}\textbf{Demonstration of low-latency spinning analysis}:
Results for marginal posterior distributions
  of our fiducial synthetic neutron star, assuming the binary has nonprecessing spins.  Solid contours show credible intervals; solid one-dimensional distributions
  show marginal CDFs and PDFs for the corresponding variable; and colored points indicate the location $\lambda$ and
  value of the underlying marginalized likelihood evaluations.  The solid curves show an analysis with SEOBNRv4\_ROM;
  the dotted curves show an analysis with SpinTaylorT4.   \emph{Left panel } Posterior distribution
  over  $\mc$,
  $\delta=(m_1-m_2)/M$, and $\chi_{\rm eff}$.    \emph{Right panel}: Marginal 1d CDFs of $\delta$, showing convergence.
\emph{Bottom left}: Mean and variance of  $\ln{\cal L}_{\rm marg}$ on the evaluation points,  showing that after the
first iteration the
candidate points are consistent with the posterior (i.e., no proposed point has very low $\ln {\cal L}_{\rm marg}$).
\emph{Bottom right panel}: $Z = \int d\lambda {\cal L}_{\rm marg}$ versus iteration number.  As systematic fitting error dominates our
error budget early on, Monte Carlo error is not shown.

}
\end{figure*}
